Follow-up Samples

### 2017-03-24 submission/manuscript_outline_20161024.tex
wrapping up mansucript text

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. This study tested the hypothesis that treatment for
adenoma or carcinoma result in changes to the bacterial community.
Specifically, we tried to identify components within the community that
were different before and after removal of lesion (adenoma or
carcinoma).

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma cases (P-value
\textless{} 0.05). Yet no difference was found in the relative abundance
of any OTU before and after treatment for adenoma or carcinoma groups
(P-value \textgreater{} 0.05). A lesion model had an AUC range of 0.692
- 0.761 and follow up samples had no difference in the positive
probability of lesion versus initial samples (P-value \textgreater{}
0.05); suggesting that the lesion associated community persists after
treatment. A treatment model had an AUC range of 0.657 - 0.796 and had a
decrease positive probability for the follow up samples to be an initial
sample (P-value \textless{} 0.05); suggesting that there are members
within the community that repond to treatment. The lesion model used a
total of 54 variables while the initial sample model used a total of 70
variables. A total of 32 OTUs were common to both models with many
classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
\emph{Ruminococcus}).

\textbf{Conclusions:} Our data partially supports the hypothesis that
the bacterial community changes after treatment. Individuals with
carcinoma have more drastic differences to the overall community then
those with adenoma. Commensal bacteria were crucial for accurate model
classification, suggesting that these bacteria may be important to
initial polyp formation and transition to carcinoma.

@@

Recent studies in humans have shown that both the bacterial community
and specific members within it correlate with CRC pathogenesis
{[}3,4{]}. Further, Dejea, et al. observed that bacterial communities
are altered between normal and tumor tissue {[}5{]}. Mouse models of CRC
have further demonstrated the importance of the microbiome, both on a
community {[}3,6{]} and species level {[}4{]}, for tumorgenesis.
Collectively, these studies provide a tantalizing link between our gut
bacteria and CRC and suggest that biomarkers using our microbes could be
developed. Indeed, builidng models using 16S rRNA gene sequencing along
with clinical tests such as Fecal Immunoglobulin Test (FIT) result in
good predictions of CRC {[}7,8{]}. Although these studies show how our
gut bacteria can impact CRC progression via a changed community or
invasion by more inflammatory bacteria {[}9{]}. They provide very little
information as to whether these changed communities rebound after
successful treatment of lesion (adenoma or carcinoma).

In this study, we tested the hypothesis that there are detectable
changes to the bacterial community between pre- (initial) and post-
(follow up) treatment of lesion. We analyzed changes in alpha and beta
diversity as well as the relative abundance of specific Operational
Taxonomic Units (OTUs). We then investigated several important genera
from oral microbes that have been suggested to be important in CRC
pathogenesis {[}9{]}. We next utilized Random Forest to build two
models: The first was built to classify lesion versus non-lesion
(normal) while the second was built to classify initial versus follow up
treatment samples. Subsequent observations on how these models and OTUs
within them performed before and after treatment helped inform us as to
whether they were changing after treatment and whether it was towards a
normal community. We also investigated the two models for similar
important OTUs to identify which were important for classifying both
lesion versus normal and initial versus follow up treatment samples.
Finally, since treatment also included chemotherapy and radiation
therapy along with lesion removal, we tested if these additionally
treatments significantly lowered any of the metrics examined versus
those who received only lesion removal. This study helps to provide
evidence as to whether treatment can influence the community and if the
CRC microbiome, identified in previous studies, persists.

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether those with adenoma (n = 41) or carcinoma (n = 26) had any
broad differences between their initial and follow up samples. We found
that those with carcinoma had a more dissimilar bacterial community
between their initial and follow up sample than those with adenoma
(P-value \textless{} 0.001) {[}Figure 1A{]}. The bacterial community
structure before and after surgery was visualized using NMDS for both
adenoma {[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05) and carcinoma
{[}Figure 1C{]} (PERMANOVA \textless{} 0.05). Interestingly, when
initial and follow up samples were compared, regardless of whether the
lesions were adenoma or carcinoma, there was no significant overall
difference in beta diversity (PERMANOVA \textgreater{} 0.05). There was
no difference between initial and follow up samples when investigating
alpha diversity metrics for lesion, adenoma only, or carcinoma only for
any metric tested {[}Table S1{]}. Additionally, there was also no
difference in the relative abundance of any OTU between initial and
follow up samples for lesion, adenoma only, or carcinoma only {[}Figure
S1{]}.

\textbf{\emph{Carcinoma Associated Genera:}} For carcinoma {[}Figure
S2A{]} (P-value \textgreater{} 0.05) but not adenoma {[}Figure S2B{]}
(P-value \textgreater{} 0.05) there was a significant decrease in
\emph{Porphyromonas}, \emph{Parvimonas}, and \emph{Peptostreptococcus}
between initial and follow up samples {[}Figure S2{]}.
\emph{Fusobacterium} for both adenoma and carcinoma had no significant
differences between initial and follow up samples {[}FIgure S2{]}
(P-value \textgreater{} 0.05). Although there were significant
differences, only a small percentage of those with adenoma or carcinoma
were positive for any of these specific genera {[}Figure S2{]}.

\textbf{\emph{The Lesion Model:}} We tested whether follow up samples
were more normal than initial samples by building a model based solely
on OTUs alone that classified lesion versus normal. This model had an
AUC range of 0.692 - 0.761 after 100 iterations of 20 repeated 10-fold
cross validations. The ROC curve for the final lesion model used was
within the observed range of the 100-different test set AUC iterations
{[}Figure 2A{]}. There was a total of 54 OTUs that were used in this
model {[}Figure 2B{]}. \emph{Lachnospiraceae} (Otu000015) resulted in
the largest decrease in MDA {[}Figure 2B{]}. Heavily represented genera
within the model included OTUs from \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
\emph{Ruminococcus}.

If there was movement towards a more normal community after treatment of
lesion, we would expect to find a decrease in the positive probability
of the follow up sample to be a lesion. We observed no such decrease in
positive probability for either those with adenoma or carcinoma
{[}Figure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
significant difference between the predicted and actual calls (P-value
\textless{} 0.05). However, this model correctly classified the one
individual who still had carcinoma on follow up and this individual's
positive probability of lesion increased between their initial and
follow up sample {[}Figure 2C{]}.

\textbf{\emph{The Treatment Model:}} To test if there were differences
that could be used to classify initial and follow up treatment samples
we built a model to classify initial samples. This treatment model had
an AUC range of 0.657 to 0.796. after 100 iterations of 20 repeated
10-fold cross validations. The test set AUC range for this model
performed better than the training set AUCs. There was a marked decrease
in the ROC curve for the final model used when compared to the 100 test
set AUC iterations {[}Figure 3A{]}. There was a total of 70 OTUs that
were used for this model {[}Figure 3B{]}. The variable that resulted in
the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
Similar to the lesion model, heavily represented genera included OTUs
from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, and \emph{Ruminococcus}.

If there were changes between initial and follow up samples we would
expect the positive probability of being an initial sample to be
decreased in the follow up. The is what we observed for the treatment
model (P-value \textless{} 0.001). When we separated lesion into adenoma
and carcinoma there was a decrease in positive probability for both the
carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
{[}Figure 3D{]} \textless{} 0.001). For this model, there was no
difference between the predicted and actual classifications (P-value
\textgreater{} 0.05).

\textbf{\emph{Common OTUs to both Models:}} We next wanted to know what
predictors within the lesion model were also in the initial sample
model. The main purpose was to identify which OTUs could be important in
lesion formation and were impacted by treatment. When we compared the
two different models with each other there were a total of 32 common
OTUs. Some of the most common taxonomic identifications belonged to
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, \emph{Anaerostipes}, \emph{Dorea}, and
\emph{Ruminococcus}. These along withthe vast majority of the OTUs that
were common between models had classifications to bacteria typically
thought of as commensal {[}Table S2{]}.

\textbf{\emph{Chemotherapy and Radiation Differences:}} After observing
these changes due to treatment we assessed whether chemotherapy or
radiation impacted the observed results. Only the treatment model had a
significant change in positive probability for those treated with
chemotherapy or radiation versus those who received neither {[}Figure 4A
\& Table S3{]} (P-value \textless{} 0.05). Additionally, there was no
difference between the change in positive probability for chemotherapy
and radiation therapy (Uncorrected P-value \textgreater{} 0.05). Using
the 32 common OTUs we tested how chemotherapy or radiation affects
important members of the community. For both chemotherapy and radiation,
the directionality of change of these OTUs were similar {[}Figure 4B \&
4C{]}. After multiple comparison correction, only \emph{Blautia}
(Otu000006) remained significantly different between those who did and
did not receive chemotherapy. This data suggests that follow up samples
from those treated with either chemotherapy or radiation may have had a
larger change from initial sample due to effects on the bacterial
community.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differ and how it might be used as an early screening tool
{[}7,8,10--12{]}. Here we describe how the bacterial community changes
after treatment and which OTUs may be most important to lesion
classification. Interestingly, many of the most important OTUs for both
treatment and lesion classification had taxonomic identification for
resident gut microbes. This suggest that members within the commensal
community may be the first that change during CRC pathogenesis. These
changes, in turn, could be the first step in allowing more inflammatory
bacteria to gain a foothold within the colon {[}9{]}.

Unlike previous studies on the microbiome and CRC, ours focuses on
identifying commonalities within both adenoma and carcinoma groups
before and after treatment. Although there were differences for genera
associated with specific bacterium linked with CRC {[}Figure S2{]}.
These changes were not consistent across lesion. Instead, most of our
results provide support that the first members of the community to
change and potentially stay changed even after treatment are those that
are commensal bacteria {[}Figure 2-3 \& Table S3{]}. Additionally,
treatment with chemotherapy or radiation changed these bacteria and
provided a larger decrease in positive probability then removal of
lesion.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are several reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from other OTUs. Third, is that
changes in commensal bacteria are the common differences across adenoma
and carcinoma. Finally, it is also possible that all these explanations
could have played a role. Our observations though, would suggest that an
individual's resident bacteria have a large role to play in polyp
formation and could change in a way that allows predictive models to
lower the positive probability of lesion or initial sample before
treatment {[}Figure 2 \& 3{]}. It should be noted that our study does
not argue against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that they are not the main bacteria
changing after treatment. It is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after treatment {[}Figure S2{]}.

Many of the common OTUs that we identified taxnomically classified to
potential butyrate producers {[}Table S2{]}. Another batch of OTUs
classified to bacteria that can either degrade polyphenols or are
inhibited by them. Both butyrate and polyphenols are thought to be
protective against cancer in part by reducing inflammation {[}13{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}14{]}. High protein consumption in general has been linked with an
increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}16{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma and carcinoma (uncorrected P-value \textless{} 0.05), with time
passed being less for adenoma (253 +/- 41.3 days) than carcinoma (351
+/- 102 days). These results would indicate that some of the differences
observed between the carcinoma and adenoma groups could be due to
differences in collection time. Specifically, it could confound the
observation that carcinomas changed more than adenomas {[}Figure 1A \&
1C{]}. This confounding though would not affect the observations where
these individuals were grouped together {[}Figures 2-4{]}.

Another limitation was that we do not know whether individuals who were
still classified as positive by the lesion model eventually had a
subsequent CRC diagnosis. This information would help to strengthen the
case for this model keeping several individuals above the cutoff
threshold even though at follow up they were diagnosed as no longer
having a lesion. This study also drew heavily from those with Caucasian
ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur.

Interestingly, within the treatment model the test data performed better
than the training data. This may have occurred because the training AUC
determined from 20 repeated 10-fold cross validation removed samples at
random and did not consider that they were matched samples. Another
potential reason is that the model itself may have been over-fit since
the total number of samples was not that large. However, the lesion
model did not suffer from these discrepancies. Further independent
studies need to be carried out to verify our findings on lesion and
treatment changes due to these limitations.

Despite these shortcomings our findings add to the existing scientific
knowledge on CRC and the microbiome: That there is a measurable
difference in the bacterial community after adenoma and carcinoma
treatment. Further, the ability for machine learning algorithms to take
OTU data and successfully lower positive probability of lesion or
initial sample after treatment provides evidence that there are specific
signatures, mostly attributable to commensal organisms, associated with
both treatment and lesion. Our data provides evidence that commensal
bacteria may be important in the development of polyps and potentially
the transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}7{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonscopy community within this time period
{[}17{]}. Our study cohort consisted of 67 individuals with an initial
sample as described and a follow up sample obtained between 188 - 546
days after treatment of lesion. This study was approved by the
University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}18{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The initial and follow up samples were sequenced on
the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}18{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}19{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on an independent data set of 423 individuals. This model was
then applied to our 67-person cohort. It should be noted that all
individuals with an adenoma or carcinoma were grouped together to form
the lesion group and the model was not created to find differences
between normal, adenoma, and carcinoma but rather differences between
both adenoma and carcinoma versus normal.

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}20{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}21{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

@@

\textbf{Figure 1: General Differences between the Adenoma and Carcinoma
Group.} A) A significant difference was found between the adenoma and
carcinoma group for thetayc (P-value = 0.000472). Advanced adenomas are
denoted as Screen Relevant Neoplasia (SRN). B) NMDS of the initial and
follow up samples for the adenoma group. C) NMDS of the initial and
follow up samples for the carcinoma group.

>>>>

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. This study tested the hypothesis that treatment for
adenoma or carcinoma results in changes to the bacterial community.
Specifically, we tried to identify components within the community that
were different before and after treatment of adenoma, advanced adenoma
(Screen Relevant Neoplasia (SRN)), and carcinoma.

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma or SRN cases
\emph{(P-value )}. But there was a trend for increasingly less community
similarity between initial and follow up samples from adenoma to SRN to
carcinoma. Despite this, no difference was found in the relative
abundance of any OTU before and after treatment for adenoma, SRN, or
carcinoma groups \emph{(P-value )}. Using Random Forest models to assess
whether changes in follow up samples were towards a normal community
showed that only those with carcinoma had a significant decrease in
positive probability \emph{(P-value)}; providing further evidence that
treatment has the greatest effect in those with carcinoma. The adenoma
model used a total of \emph{model n} OTUs, the SRN model used a total of
\emph{model n} OTUs, and the carcinoma model used a total of \emph{model
n} OTUs. A total of \emph{common n} OTUs were common to both models with
many classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
\emph{Dorea}). Both chemotherapy and radiation did not provide any
additional changes to the bacterial community in those treated for
carcinoma \emph{(Pvalue )}

\textbf{Conclusions:} Our data partially supports the hypothesis that
the bacterial community changes after treatment. Individuals with
carcinoma had more drastic differences to the overall community then
those with adenoma. Common OTUs to all models were overwhelmingly from
commensal bacteria, suggesting that these bacteria may be important to
initial polyp formation, development of advanced adenoma, and transition
to carcinoma.

@@

Recent studies in humans have shown that both the microbiome and
specific members within it correlate with CRC pathogenesis {[}3,4{]}.
Further, bacterial communities have been observed to be altered between
normal and tumor tissue {[}5{]}. Mouse models of CRC have further
demonstrated the importance of the microbiome, both on a community
{[}3,6{]} and species level {[}4{]}, for tumorgenesis. Collectively,
these studies provide a tantalizing link between our gut bacteria and
CRC and suggest that biomarkers using our microbes could be developed.
Indeed, builidng models using 16S rRNA gene sequencing along with
clinical tests such as Fecal Immunoglobulin Test (FIT) result in good
predictions of CRC {[}7,8{]}. Although these studies show how our gut
bacteria can impact CRC progression via a changed community or invasion
by more inflammatory bacteria {[}9{]}. They provide very little
information as to whether these communities change and rebound towards
normal after successful treatment of adenoma, advanced adenoma (Screen
Relevent Neoplasia (SRN)), or carcinoma.

Providing an answer to this question is important because it has far
reaching implications on both how the bacterial community causes the
formation of more polyps {[}3,4{]} and the ability to be able to use the
microbiome as a predictive screening tool {[}7,8{]}. Understanding polyp
formation and transition to SRN and then carinoma is crucial to being
able to understand how to prevent CRC occurence. Response of the
community to treatment is also eqaully important to predictive models
designed for screening purposes since an unresponsive community would
provide little additional information for important events, such as
recurrence {[}10{]}.

Using pre- (initial) and post- (follow up) treatment samples we tested
the hypothesis that treatment causes detectable changes to the
microbiome in those with adenoma, SRN, and carincoma. First, we assessed
differences between initial and follow up samples in adenoma, SRN, or
carinoma using alpha or beta diversity metrics. Second, we explored
whether models built to classify adenoma, SRN, or carcinoma versus
normal were able to identify specific community members that differed
between initial and follow up. We also used these models to assess
whether changes in the community were toward a more normal micorbiome.
Finally, we assessed both whether surgery for adenomas and SRN provided
larger community changes or whether chemotherapy or radiation provided
additive changes to the microbiome over surgerical resection. This study
helps to provide evidence as to whether treatment can influence the
community and if the CRC microbiome, identified in previous studies,
persists after such interventions.

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether those with adenoma \emph{(n = )}, SRN \emph{(n = )}, or
carcinoma \emph{(n = )} had any broad differences between their initial
and follow up samples. We found that those with carcinoma had a more
dissimilar bacterial community between their initial and follow up
sample than those with adenoma \emph{(P-value 0.001)} {[}Figure 1A{]}.
Although no significant differences were observed between SRN and
carcinoma there was an increase in the dissimilarity of the initial and
follow up samples from adenoma \emph{(mean and SD)} to SRN \emph{(mean
and SD)} to carcinoma \emph{(mean and SD)} {[}Figure 1A{]}. The
bacterial community structure before and after surgery was visualized
using NMDS for adenoma {[}Figure 1B{]} \emph{(PERMANOVA 0.05)}, SRN
{[}FIgure 1C{]} \emph{(PERMANOVA 0.05)}, and carcinoma {[}Figure 1C{]}
\emph{(PERMANOVA 0.05)}. Interestingly, when initial and follow up
samples were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
\emph{(PERMANOVA 0.05)}. There was no difference between initial and
follow up samples when investigating alpha diversity metrics for
adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
Additionally, there was also no difference in the relative abundance of
any OTU between initial and follow up samples for adenoma, SRN, or
carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
20 repeated 10 fold cross-validation was \emph{from to this} with the
AUC of the model used being \emph{AUC value} {[}Figure S2A{]}. There was
a total of \emph{blank} OTUs in this model with the vast majority
classifying to bacteria typically thought of as commensal {[}Figure
S3A{]}. There was a significant difference between the actual and model
predicted group stratification calls \emph{(P-value)}. There was no
significant decrease in the positive probability of adenoma between
initial and follow up samples \emph{(P\_value)} {[}Figure 2A{]}.

\textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
repeated 10 fold cross-validation was \emph{from to this} with the AUC
of the model used being \emph{AUC value} {[}Figure S2B{]}. There was a
total of \emph{blank} OTUs in the SRN model {[}Figure S3B{]}. Similar to
the adenoma model the vast majority of OTUs classified to bacteria
typically thought of as commensal. Also similar to the adenoma model
there was a significant difference between the actual and model
predicted group stratification calls \emph{(P-value)}. There was no
significant decrease in the positive probability of SRN between initial
and follow up samples \emph{(P\_value)} {[}Figure 2B{]}.

\textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
of 20 repeated 10 fold cross-validation was \emph{from to this} with the
AUC of the model used being \emph{AUC value} {[}Figure S2C{]}.
Interestingly, the AUCs improved from adenoma to SRN to carcinoma
{[}Figure 2{]}. There was a total of \emph{blank} OTUs in the carinoma
model {[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
majority of OTUs classified to bacteria typically thought of as
commensal but OTUs that also classified to \emph{Fusobacterium},
\emph{Porphyromonas}, and \emph{Parvimonas} appear to be important for
carcinoma classification {[}Figure S3C{]}. Also similar to the adenoma
and SRN models there was a significant difference between the actual and
model predicted group stratification calls \emph{(P-value)}. There was a
significant decrease in the positive probability of carcinoma between
initial and follow up samples \emph{(P\_value)} {[}Figure 2C{]};
suggesting that the carcinoma samples changed towards normal after
treatment, unlike either adenoma or SRN. The one indivdiual still
positive for carcinoma after treatment had an increase in carcinoma
positive proability on follow up {[}Figure 2C{]}.

\textbf{\emph{Adenoma, SRN, and Carcinoma Common OTUs}} We next wanted
to know what predictors within the adenoma, SRN, and carcinoma models
were similar to each other. The main purpose was to identify which OTUs
could be important at all three stages of disease. When we compared the
three different models with each other there were a total of
\emph{number} common OTUs. Some of the most common taxonomic
identifications belonged to \emph{one}, \emph{two}, \emph{three},
\emph{four}, \emph{five}, \emph{six}, and \emph{seven}. These along with
the vast majority of the OTUs that were common between these models had
classifications to bacteria typically thought of as commensal {[}Table
S2{]}.

\textbf{\emph{Treatment Affects on Community}} After observing these
these changes from treatment we assessed whether chemotherapy or
radiation, in the carcinoma group, and surgery, in the adenoma group,
impacted the observed results. In the carcinoma group neither
chemotherapy or radiation provided any additive change from initial
sample over surgery alone {[}Table S3{]}. For the adenoma group there
was no difference between those that received surgery and those that did
not {[}Table S4{]}. For surgery adenoma and SRN were combined due to the
low number of surgery occurances. There was no difference in the
proportion of those receiving surgery between the adenoma and SRN groups
\emph{(P-value)}. This data suggests that microbiome changes observed in
the carcinoma group were mostly a result from surgery and not from
chemotherapy or radiation.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differ and how it might be used as an early screening tool
{[}7,8,11--13{]}. Here we show that the bacterial community changes
towards normal after treatment for carcinoma and that chemotherapy and
radiation did not provide an additive change. Although some of the
important OTUs classified to genera from bacteria considered the usual
suspects (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas}) many did not. The majority of important OTUs had
taxonomic classifications for resident gut microbes and were common for
the adenoma, SRN, and carcinoma models. This suggests that members
within the commensal community may be the first that change during CRC
pathogenesis. These subtle changes, in turn, could be the first step in
allowing more inflammatory bacteria to gain a foothold within the colon
{[}9{]}.

Unlike previous studies on the microbiome and CRC, ours focuses both on
identifying commonalities and differences within adenoma, SRN, and
carcinoma groups before and after treatment. Although there were
differences for genera associated with specific bacterium linked with
CRC {[}Figure S3{]}. The majority of important OTUs taxonmically
classified to commensal bacteria {[}Figure S3{]}. Although these changes
may be subtle, due to the lack of signifcant difference in the bacterial
community before and after treatment in adenoma and SRN {[}Figure 1{]},
they support the hypothesis that the first members of the community to
change and potentially stay changed even after treatment are those that
are commensal bacteria.

Many of the common OTUs that we identified taxnomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}, and \emph{Anaerostipes}){[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}14{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}15{]}. High protein consumption in general has been linked with an
increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}17{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma or SRN versus carcinoma \emph{(uncorrected P-value 0.05)}, with
time passed being less for adenoma \emph{( days)} and SRN \emph{( days)}
than carcinoma \emph{( days)}. These results would indicate that some of
the differences observed between the carcinoma and adenoma groups could
be due to differences in collection time. Specifically, it could
confound the observation that carcinomas changed more than adenomas
{[}Figure 1A \& 1C{]}. However, there are two reasons that this may not
be the case. First, the SRN group did not have a significant difference
and the collection time of their follow up sample was less than the
adenoma group. Second, this confounding would not affect the
observations where models were used since they were built using a
different cohort {[}Figures 2 \& S2-S3{]}.

Another limitation was that we do not know whether individuals who were
still classified as positive by the carcinoma model eventually had a
subsequent CRC recurrence. This information would help to strengthen the
case for this model keeping numerous individuals above the cutoff
threshold even though at follow up they were diagnosed as no longer
having carcinoma. This study also drew heavily from those with Caucasian
ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur.

Despite these shortcomings our findings add to the existing scientific
knowledge on CRC and the microbiome: That there is a measurable
difference in the bacterial community after adenoma, SRN or carcinoma
treatment. Further, the ability for machine learning algorithms to take
OTU data and successfully lower positive probability of carcinoma after
treatment provides evidence that there are specific signatures,
attributable to both inflammtory and resident commensal organisms,
associated with treatment. Our data provides evidence that commensal
bacteria may be important in the development of polyps and potentially
the transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}7{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonscopy community within this time period
{[}18{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}19{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The initial and follow up samples were sequenced on
the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}19{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}20{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on an independent data set of 423 individuals. This model was
then applied to our 67-person cohort. It should be noted that all
individuals with an adenoma or carcinoma were grouped together to form
the lesion group and the model was not created to find differences
between normal, adenoma, and carcinoma but rather differences between
both adenoma and carcinoma versus normal.

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}21{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}22{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

@@

\textbf{Figure 1: General Differences between Adenoma, SRN, and
Carcinoma Groups After Treatment.} A) A significant difference was found
between the adenoma and carcinoma group for thetayc (P-value = NULL).
Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN). B)
NMDS of the initial and follow up samples for the adenoma group. C) NMDS
of the initial and follow up samples for the carcinoma group.

\textbf{Figure 2: Treatment Response Based on Models Built for Adenoma,
SRN, or Carcinoma.} A) Positive probability change from initial to
follow up sample in those with adenoma. B) Positive probability change
from initial to follow up sample in those with SRN. C) Positive
probability change from initial to follow up sample in those with
carcinoma..

\textbf{Figure 3: The Treatment Model.} B) C) Positive probability
change from initial to follow up sample in those with carcinoma. D)
Positive probability change from initial to follow up sample of those
with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
